Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-09-21 paper/manuscript.tex
Updates to free energies of hydration/ partition coefficients by @megosato in response to review
Changes to uncertainty quantification sections from @hbruce-beep while reviewing

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016}. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive allowing one to reduce the computational costs when resources are limited~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities. Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.

@@

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
% Add something on RBFE calculations across engines.
% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
% - other references?

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

\subsection{Free energies of binding}
\label{subsec:binding}
In principle, in the limit of sufficient configurational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice, because simulations are of finite duration (typically 1-100 ns per state at present), this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that significantly exceed the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
The following questions may be relevant before diving into the simulation setup.

@@

\paragraph{The choice of solvent force field impacts the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. In a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.

\paragraph{Conformational sampling can impact solvation and partition free energies}
Moreover, while sampling challenges are often perceived to be minor for small molecules, they can still be non-trivial. Even subtle degrees of freedom such as hydroxyl group orientations (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.

@@

\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future. 
Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.

@@

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Replicas in ensemble simulations are initialized with velocities drawn independently from the Maxwell–Boltzmann distribution and hence generate decorrelated samples by default. Ensemble approaches are therefore strongly preferred for uncertainty estimation, as single simulations remain constrained by the timescales they can access and may therefore underestimate the true uncertainty. Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}. When only one simulation is available, the procedures described below provide a lower-bound estimate of the uncertainty.

@@

Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be ``ergodic'' but it must also be ``mixing'' (these terms refer to the different classes of dynamical systems provided by ergodic theory; please see the references for details)~\cite{coveney2016, MDbook2025}. Rate processes (a course of change within a system that occurs over time) and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral. It assumes that the integrand is a Gaussian random process and hence the thermodynamic integral can be viewed as a so-called stochastic integral, which forms the basis of uncertainty propagation in this integral. The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state. $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.

@@

Compared with TI, the MBAR method (Sec.~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states. 
The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
%A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost. 
However, this process based on one-off simulations underestimates the true uncertainty and hence uncertainties for MBAR also should be obtained using ensemble simulations~\cite{bhati2018, wade2022}.

@@

In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations. In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices. Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}. They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.

>>>>

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025}. Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.

@@

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
% Add something on RBFE calculations across engines.
% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
% - other references?

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of hydration or a free energy of binding. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.

\subsection{Free energies of binding}
\label{subsec:binding}
In principle, in the limit of sufficient configurational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice, because simulations are of finite duration (typically 1-100 ns per state at present), this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that significantly exceed the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. The above considerations in Section \ref{subsec:hydration} on preparing solvation free energy and partition coefficient calculations also apply to binding free energy calculations. 
In addition to the above, the following questions may be relevant before diving into the binding free energy simulation setup.

@@

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative, bhati2017, bieniek2021} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
%TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.

@@

\paragraph{Conformational sampling can also solvation and binding free energies}
While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial. Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

These challenges are magnified in binding free energy calculations, where sampling all relevant conformations within reasonable timescales becomes even more difficult.

\subsubsection{What $\lambda$-sampling schemes are available?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

%Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future. 
%Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.

@@

\subsubsection{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

@@

\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Replicas in ensemble simulations are initialized with velocities drawn independently from the Maxwell–Boltzmann distribution and hence generate decorrelated samples by default. When only one simulation is available, the procedures described below provide a lower-bound estimate of the uncertainty.

@@

%Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be ``ergodic'' but it must also be ``mixing'' (these terms refer to the different classes of dynamical systems provided by ergodic theory; please see the references for details)~\cite{coveney2016, MDbook2025}. Rate processes (a course of change within a system that occurs over time) and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

%The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral. It assumes that the integrand is a Gaussian random process and hence the thermodynamic integral can be viewed as a so-called stochastic integral, which forms the basis of uncertainty propagation in this integral. The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state. $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.

Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propogation.

@@

Compared with TI, the MBAR method (Sec.~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states. 
The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost. 
However, this process based on one-off simulations underestimates the true uncertainty and hence uncertainties for MBAR also should be obtained using ensemble simulations~\cite{bhati2018, wade2022}.

@@

In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations. In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices. Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
Incidentally, Vassaux \textit{et al.} perfor med a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}. They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
